# Introduction
A common problem in data science is applying a function to a set of rows grouped by some feature using the `group_by()` and `summarise()` workflow. But what if one needs to collapse the data by some summary measure then join the summary measure back into the original dataset?

The generalized problem can be described as:

"For each id, perform some operation on or return a value from column x, given a condition in column x, or possibly column y"

An overall workflow to consider is:

1. `group_by()`
2. Extract summary data values by creating a temporary variable
3. `mutate()`

Below we show some workflows for how to tackle specific "row-oriented" problems. 

We start by creating a sample data set for patient groups A-F. For simplicity, there are three patients within each group.
```{r}
# Load packages
library(dplyr)

# Create a sample data for different patient groups
my_df <- data.frame(patient_id = 1:18,
                    patient_grp = factor(rep(LETTERS[1:6], each=3)),
                    a = rep(1:2),
                    b = c(11:13, 21:23, 31:33, 41:43, 51:53, 61:63),
                    x = c(1:6, 1:12),
                    y = c(rep("aa", 6), rep("bb", 12)))

# Create a tibble
my_tibb <- my_df %>% 
  as_tibble()
my_tibb
```

### Counting within group and returning the most frequent

In the first example, we want to count within groups.

Here the `dplyr` function `add_count` comes in handy. Note that `add_count` is a shortcut step for `group_by()` %>% `add_tally()`. 


```{r}

my_tibb %>%
  # add a column n with count by categories
  add_count(patient_grp, a) %>%
  # select max or first occurrence by patient
  group_by(patient_grp) %>%
  # keep only first TRUE
  mutate(majority = a[n == max(n)][1]) %>%
  # do not keep temp var
  select(-n)
```

### The use of `case_when` to sort within a group

In this example, we want to `sort()` within group. In this example, we want to sort the data frame by `b`.  However, we want to sort groups "C" and "F" decreasing and the remaining groups increasing. 

The overall workflow is to first sort with the flag `decreasing = TRUE` then use the `dplyr` function `case_when` to selectively sort

This could be generalized to any function, not just `sort()`.
```{r}
# The desired vector looks like this
desired <- c(11:13, 21:23, 33:31, 41:43, 51:53, 63:61)

# i.e. when the patient group is C, the order of column b is reversed

my_tibb %>%
  group_by(patient_grp) %>%
  mutate(b_sort_dec = sort(b, decreasing = TRUE)) %>%
  mutate(b_desired = case_when(patient_grp %in% c("C", "F") ~ b_sort_dec,
                              TRUE ~ b))
```

You can then remove the `b_sort_dec` columns you don't want to keep

### Filtering columns by column pairs

Another situation is when you want to filter a data frame by x-y pairs to only keep rows where there is an x and y for each. For example, if there are two possible values `aa` and `bb` we only want to keep those rows whichh as an `x` = `aa` AND an `x` = `bb`. In other words, we want to remove all rows where we have an`x` = `aa` but no corresponding `x` = `bb`, or vice versa.

In our example data frame, we have 18 rows. However, only 12 of these rows have x-y pairs such that there is an `x` = `aa` and an `x` = `bb`. We effectively want to remove the final 6 rows. Again, the dplyr function `add_count` comes in handy.


```{r}
my_tibb %>%
  add_count(x) %>% 
  filter(n > 1)

```


# Further reading

* Link to Jenny Bryan's row-oriented data manipulations

Like SQL, dplyr uses window functions that are used to subset data within a group. An understanding of SQL window functions could be helpful.

* Link to SQL Window functions

* Romain's rap package


